---
title: "Immune Callenge Assay"
author: "Dylan Ryals"
date: "2024-04-03"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Results from:

Why diversity matters: non-additive fitness benefits of individual versus colony-level genetic variation in social insects

by Ryals, D.K., Buschkoetter, A.C., Given, K.J., and Harpur, B.A.


Honey bee immune challenge assay, summer 2023.


## Setup 

load packages and read data

```{r}
library(tidyverse)
library(ggrepel)
library(readxl)
library(lubridate)
library(survival)
library(survminer)

select = dplyr::select
mutate = dplyr::mutate

### Read in, clean data

assay = read_excel("immune_data/observations.xlsx", sheet = "removed",
                   col_types = c("date", "numeric", "text", "numeric", 
                                 "numeric", "text", "text")) %>%
        suppressWarnings()

assay$cup = as.character(assay$cup)

cups = read_excel("immune_data/observations.xlsx", sheet = "cups",
                  col_types = c("numeric", "date", "date", "text", "numeric", "numeric",
                                "text", "text"))%>%
        suppressWarnings()


cups$cup = as.character(cups$cup)
cups$id = as.character(cups$id)
cups$color = toupper(cups$color)

key = read_excel("immune_data/id_key.xlsx", sheet = 'Sheet1')
key$id = as.character(key$id)
key$rep = as.character(key$rep)

cups = cups %>% left_join(key)

```


Format and clean data for survival analysis

```{r}
#define mixture for mixed micro colonies
cuptype = data.frame(cup = unique(cups$cup), mix = NA)
for(i in 1:nrow(cuptype)){
  mix = cups$cross[cups$cup == cuptype$cup[i]] %>% sort(decreasing = T)
  if(length(mix) == 2){
    cuptype$mix[i] = paste0(mix[1], "+", mix[2])
  }
  else cuptype$mix[i] = mix
}

cups = cups %>% left_join(cuptype, by = 'cup')

#see cup totals
summary = assay %>% filter(!is.na(color)) %>%
  group_by(cup, color) %>% 
  summarise(sum_dead = sum(count, na.rm = T))


#convert to long
assay.narm = assay %>% filter(!is.na(count))
assay.long = data.frame(datetime = NA, cup = NA, color = NA, status = NA)

for(i in 1:nrow(assay.narm)){
  #select relevant data in ith observation
  tmp = assay.narm[i,] %>% select(-count, -observer, -note)
  #repeat observation for each count (one row per bee removed)
  for(j in 1:assay.narm$count[i]){
    assay.long = rbind(assay.long, tmp)
  }
}
assay.long = assay.long[-1,]


assay2 = assay.long %>% 
  left_join(cups %>% select(cup, color, id, control, start, run, cross, mix, rep), 
            by = join_by(cup, color))

assay2$start = as.numeric(assay2$start)
assay2$runtime = assay2$datetime - assay2$start

#fix order of one mix name
assay2$mix[assay2$mix == "AB+AB"] = "AB"

#convert seconds to days
assay2$days = assay2$runtime / (60 * 60 * 24)

#remove cups with irregular counts 
assay2 = assay2 %>% filter(! cup %in% c(24, 26, 31))
#remove configuration trials
assay2 = assay2 %>% filter(cup > 18)


### three category: inbred, cross, and mix
test3 = assay2 %>% filter(color != "?", 
                          mix %in% c('BB', 'AA', 'BB+AA', 'BA', 'AB'))

test3$cat = NA
test3$cat[test3$mix %in% c("BB", "AA")] = "inbred"
test3$cat[test3$mix %in% c("AB", "BA")] = "hybrid"
test3$cat[test3$mix == "BB+AA"] = "mixed"


##TODO
  #double-check censored data...
```


## Survival Analysis

```{r}

#treatment different from control?
survdiff(Surv(days, status)~control + strata(cat), data = test3)

#comparing treatments
test4 = test3
test4$cat[test4$control == 1] = "control"
test4$cat = factor(test4$cat, levels = c("inbred", "hybrid", "mixed", "control"))

#main figure
group.colors = c("#648FFF", "#FFB000", "#DC267F")
group.colors2 = c("#648FFF", "#FFB000", "#DC267F", "gray50")

sfit3 = survfit(Surv(days, status)~cat, data = test4)

gsp = ggsurvplot(sfit3, conf.int = T, palette = group.colors2) + 
  labs(x = "Days")
gsp

#save figure
ggsave("figures/figure2_immune.pdf", device = "pdf", units = "in",
       width = 10.25, height = 7.35)
  #figure imported into vector graphic editor for small edits

#number of
  #trials
  test4 %>% filter(control == 0) %>% summarise(ncups = length(unique(cup)))
  #controls
  test4 %>% filter(control == 1) %>% summarise(ncups = length(unique(cup)))
  #total days
  max(test4$days)
  #total days control
  test4 %>% filter(control == 1) %>% summarise(max(days))
  #total deaths in control
  test4 %>% filter(control == 1) %>% summarise(sum(status == 1))
  #total deaths in trials
  test4 %>% filter(control == 0) %>% summarise(sum(status == 1))
  #total trials in each replicate
  test4 %>% filter(control == 0) %>% 
    group_by(rep, cat) %>% 
    summarise(length(unique(cup)))

  

###test significance with log-rank tests
  lr.mi = survdiff(Surv(days, status)~cat + strata(rep, cross), data = test4, 
                   subset= cat %in% c("mixed","inbred"))
  
  lr.hi = survdiff(Surv(days, status)~cat + strata(rep), data = test4, 
                   subset= cat %in% c("hybrid","inbred"))
  
  lr.hm = survdiff(Surv(days, status)~cat + strata(rep), data = test4, 
                   subset= cat %in% c("mixed","hybrid"))
  
  
  #p values:
  logrank = data.frame(comparison = c("inbred:hybrid", "inbred:mixed", "hybrid:mixed"),
                    Pval = c(lr.hi$pvalue, lr.mi$pvalue, lr.hm$pvalue))
  
  logrank

  
### and cox proportional hazard model
  
cox1 = coxph(Surv(days, status)~ cat + strata(rep), data = test4)
summary(cox1)
```


## Supplementary Figure

```{r}
###crosses and partners
  
  same = assay2$cross == assay2$mix
  #remove controls
  indiv = assay2 %>% filter(control == 0)
  
  #add partners
  indiv$partner = NA
  mix = str_split_fixed(indiv$mix, "[+]", 2)
  for(i in 1:nrow(indiv)){
    if(mix[i,2] == ""){indiv$partner[i] = indiv$cross[i]}
    else if(mix[i,1] == indiv$cross[i]){indiv$partner[i] = mix[i,2]}
    else{indiv$partner[i] = mix[i,1]}
  }
  
  
sfit4 = survfit(Surv(days, status) ~ cross + partner, data = indiv)
  
ggsurvplot_facet(fit = sfit4, data = indiv, facet.by = 'cross', conf.int = T, pval = T)

#save figure
ggsave("figures/supplemental3_immune-mixes.pdf", device = "pdf", units = "in",
       width = 6.38, height = 6.54)

```
