---
title: "Thermoregulation Assay"
author: "Dylan Ryals"
date: "2024-03-28"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Results from:

Why diversity matters: non-additive fitness benefits of individual versus colony-level genetic variation in social insects

by Ryals, D.K., Buschkoetter, A.C., Given, K.J., and Harpur, B.A.


Honey bee thermoregulation experiment, summer 2023.


## Setup 

```{r read data}
#load required libraries
#####

library(tidyverse)
library(scales)
library(ggrepel)
library(readxl)
library(lubridate)
library(knitr)
library(ggsignif)
library(cowplot)
library(rmarkdown)

#custom standard error function
std.error <- function(x) sd(x)/sqrt(length(x))

#set gg theme
theme_set(theme_bw())


#read temperature sensor (iButton) data
#####

files = list.files("therm_data/sensors")

ib = data.frame(Date = NA, Time = NA, Value = NA, serial = NA)

for (f in files){
  path = paste0("therm_data/sensors/", f)
  #read
  suppressWarnings({tmp = read_excel(path, skip = 21)})
  #strip last line
  tmp = tmp[-(nrow(tmp)),]
  #grab name
  tmp$serial = gsub("(^[^_]*)([_].*$)", "\\1", f)
  #map
  ib = rbind(ib, tmp)
}
#remove first row
ib = ib[-1,]

#format cols
ib$temp = as.numeric(ib$Value)
ib$datetime = as_datetime(paste(ib$Date, ib$Time))

#read IB key to attach serial nubers to experimental groups
ibkey = read_xlsx("therm_data/ibkey.xlsx")
ibkey$serial_3 = as.character(ibkey$serial_3)
ibkey$serial = paste0(ibkey$serial_1, ibkey$serial_2, 
                      ibkey$serial_3, ibkey$serial_4)
ibkey$rep = as.factor(ibkey$rep)
ibkey$assay_id = as.factor(ibkey$assay_id)

#attach experimental data
ib = ib %>% left_join(ibkey %>% select(serial, assay_id), by = 'serial')

#preview dataframe
ib.preview = ib %>% 
  group_by(assay_id) %>% 
  arrange(datetime) %>%
  slice(1:10)

paged_table(ib.preview)
```


## Intro plots

thermoregulation assay raw results. Sensor data without trimming to desired time-frame 

```{r}
#therm results
ggplot(data = ib %>% filter(assay_id != 15)) +
  geom_line(aes(x = datetime, y = temp, color = assay_id)) +
  geom_line(data = ib %>% filter(assay_id == 15),
              aes(x = datetime, y = temp), color= 'black') +
  theme(legend.position = "none")
```



## Response variables

Trim time period and calculate response variables for each nucleus colony:

- hourly divergence from optimal temperature
- hourly variance in temperature

```{r}
#trim data to desired 4-day period

start.ib = as.POSIXct("2023-08-30 15:00:00", tz="UTC")
end.ib = as.POSIXct("2023-09-03 15:00:00", tz="UTC")

t_total = as.numeric((end.ib - start.ib), units = "hours")

ib.cut = ib %>% filter(datetime > start.ib, 
                       datetime < end.ib)
###
#therm stats

#optimal temperature
TNEST = 34.5

# 1-hr windows
ib.win = ib.cut

windows = 
  seq(
    from=start.ib,
    to= end.ib,
    by="hour"
  )

#data frame for variables
ib.variables = data.frame(var = NA, err = NA, mtemp = NA, hour = NA, assay_id = NA)

#loop through each sensor
for( i in unique(ib.win$assay_id)){
  
  #loop through each 1-hr window
  tvar = data.frame(var = rep(NA, length(windows)-1))
  
  for(j in 1:(length(windows)-1)){
    tmp.data = ib.win %>% filter(assay_id == i, datetime > windows[j],
                        datetime <= windows[j+1]) %>% select(temp)
    #thermal variance
    tvar$var[j] = var(tmp.data$temp)
    #thermal divergence (average across 12 datapoints)
    tvar$err[j] = sum(abs(tmp.data$temp - TNEST))/12
    tvar$mtemp[j] = mean(tmp.data$temp, na.rm = T)
    tvar$hour[j] = windows[j+1]
    tvar$assay_id[j] = i
  }
  #map output
  ib.variables = rbind(ib.variables, tvar)
}
#remove first NA row
ib.variables = ib.variables[-1,]
#convert to numeric hour
ib.variables$hour = as_datetime(ib.variables$hour, tz = "UTC")

#summarise mean and se for each colony
ib.variables.sum = ib.variables %>% group_by(assay_id) %>% 
  summarise(meanvar = mean(var),
            sevar = std.error(var),
            meanerr = mean(err),
            seerr = std.error(err))

#add keys for experimental groups
therm = ib.variables.sum %>% 
  left_join(ibkey %>% select(assay_id, treatment, rep), by = 'assay_id')

paged_table(therm)
```


visualize trimmed data

```{r}
ib.cut2 = ib.cut
#add control
ib.cut2$control = ifelse(ib.cut2$assay_id == 15, "ambient", "nuclei")
#add dummy horizontal line at optimal temperature
  ib.add = ib.cut2 %>% filter(assay_id == 15) %>%
    mutate(temp = TNEST, control = "optimal (35.4 C)", assay_id = 16)
  ib.cut2 = rbind(ib.cut2, ib.add)
ib.cut2$control = factor(ib.cut2$control, levels = c("nuclei", "optimal (35.4 C)", "ambient"))

#add experimental groups
ib.cut2 = ib.cut2 %>% 
  left_join(ibkey %>% select(assay_id, treatment))
  #refactor
  ib.cut2$treatment[ib.cut2$treatment %in% c("AA", "BB")] = "inbred"
  ib.cut2$treatment[ib.cut2$treatment %in% c("BA", "AB")] = "hybrid"
  ib.cut2$treatment[ib.cut2$treatment %in% c("AA+BB")] = "mixed"
  ib.cut2$treatment[ib.cut2$treatment =="control"] = NA
  ib.cut2$treatment = factor(ib.cut2$treatment, levels = c("inbred", "hybrid", "mixed"))

group.colors = c("#648FFF", "#FFB000", "#DC267F")
  
therm.temps = 
ggplot(data = ib.cut2) +
  geom_line(aes(x = datetime, y = temp, color = treatment, group = assay_id,
                linetype = control), size = 0.6, alpha = 0.8) +
  scale_color_manual(values = group.colors, na.value = "black") +
  scale_linetype_manual(values= c(1,2,4)) +
  labs(color = NULL, y = 'Temperature (\u00B0C)', linetype = NULL, x = NULL) +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 11))
  
therm.temps
```


## Statistical Analysis

Each varaiable is analysed using a paried t test, pairing observations within replicates.


```{r}
#create categories (experimental groups)
therm$cat = NA
  therm$cat[therm$treatment == "AA" | therm$treatment == "BB"] = "inbred"
  therm$cat[therm$treatment == "BA" | therm$treatment == "AB"] = "hybrid"
  therm$cat[therm$treatment == "AA+BB"] = "mixed"
  therm$cat = factor(therm$cat, levels = c("inbred", "hybrid", "mixed"))

  #remove control colony
  therm2 = therm %>% filter(!is.na(cat)) %>% 
    arrange(rep)
   therm2$unique_id = 1:nrow(therm2)
  
  
#calculate means within categories
  therm.allmeans = therm2 %>% group_by(cat) %>%
    summarise(mean_allvar = mean(meanvar), mean_allerr = mean(meanerr))
#calculate means within reps
  therm.means = therm2 %>% group_by(cat, rep) %>%
    summarise(mean_repvar = mean(meanvar), mean_reperr = mean(meanerr))

paged_table(therm.means)

#paired t tests: take means for both variables, convert to wide format
paired = therm %>% group_by(rep) %>%
  #mean of inbred lines
  filter(treatment %in% c("AA", "BB")) %>%
  summarise(inbred_var = mean(meanvar), inbred_err = mean(meanerr)) %>%
  #mixed group
  left_join(therm %>% 
              filter(treatment == "AA+BB") %>% 
              select(mix_var = meanvar, mix_err = meanerr, rep)) %>%
  #mean of hybrid lines
  left_join(therm %>% 
              filter(treatment %in% c("AB", "BA")) %>% 
              group_by(rep) %>% 
              summarise(hybrid_var = mean(meanvar), hybrid_err = mean(meanerr)))

#pvalues from paired t-tests
  #thermal error: hybrid, mix, hybrid vs mix
  test.ih.err = t.test(Pair(inbred_err, hybrid_err)~1, data = paired, alternative = "greater")
    test.ih.err
  test.im.err = t.test(Pair(inbred_err, mix_err)~1, data = paired, alternative = "greater")
    test.im.err
  test.hm.err = t.test(Pair(hybrid_err, mix_err)~1, data = paired, alternative = "greater")
    test.hm.err
    
  err.annotate = c(test.ih.err$p.value, test.im.err$p.value, test.hm.err$p.value) %>%
    round(digits = 4) %>% as.character()
  
  #thermal variance: hybrid, mix, hybrid vs mix
  test.ih.var = t.test(Pair(inbred_var, hybrid_var)~1, data = paired, alternative = "greater")
    test.ih.var
  test.im.var = t.test(Pair(inbred_var, mix_var)~1, data = paired, alternative = "greater")
    test.im.var
  test.hm.var = t.test(Pair(hybrid_var, mix_var)~1, data = paired, alternative = "greater")
    test.hm.var
    
  var.annotate = c(test.ih.var$p.value, test.im.var$p.value, test.hm.var$p.value) %>%
    round(digits = 4) %>% as.character()
```

## Main Figure

```{r}
#therm err
p.err = ggplot(therm2, aes(x = cat, y = meanerr)) +
  #bars for means
  geom_col(data = therm.allmeans, aes(y = mean_allerr, fill = cat),
           position = "dodge", alpha = 0.5, width = 0.5,
           show.legend = F) + 
  #individual points
  geom_pointrange(aes(ymax = meanerr+seerr, ymin = meanerr-seerr, shape = rep),
                  position=position_dodge(width=0.3)) +
  scale_fill_manual(values = group.colors) +
    #text and significance
  labs(y = "Hourly Error (\u00B0C)",  x = NULL, shape = "replicate") +
    geom_signif( aes(color = NULL), step_increase = 0.15,
    comparisons = list(c("inbred", "hybrid"), c("inbred", "mixed"), c("hybrid", "mixed")),
    annotations = err.annotate, textsize = 3) +
  theme(axis.text.x.bottom = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank())
  
  #therm var
p.var = ggplot(therm2, aes(x = cat, y = meanvar)) +
  #bars for means
  geom_col(data = therm.allmeans, aes(y = mean_allvar, fill = cat),
           position = "dodge", alpha = 0.5, width = 0.5,
           show.legend = F) + 
  #individual points
  geom_pointrange(aes(ymax = meanvar+sevar, ymin = meanvar-sevar, shape = rep),
                  position=position_dodge(width=0.3)) +
  scale_fill_manual(values = group.colors) +
    #text and significance
  labs(y = "Hourly Variance (\u00B0C^2)", shape = "replicate", x = NULL) +
    geom_signif( aes(color = NULL), step_increase = 0.15,
    comparisons = list(c("inbred", "hybrid"), c("inbred", "mixed"), c("hybrid", "mixed")),
    annotations = var.annotate, textsize = 3) +
    theme(panel.border = element_blank(),
          axis.text.x = element_text(size = 11))
  
  #manuscript figure
both.vars = plot_grid(p.err, p.var, labels = c("", "C"), nrow = 2)

plot_grid(therm.temps, both.vars, labels = "AUTO", ncol = 2,
          rel_widths = c(1, 0.6))

#save figure
ggsave("figures/figure3_therm.pdf", device = "pdf", units = "in",
       width = 11.74, height = 5.81)
  #figure imported into vector graphic editor for small edits
```


## Supplemental Figures

calculation of variables 
```{r}
#examples of error 
varstart = as.POSIXct("2023-08-31 20:00:00", tz="UTC")
varstop = as.POSIXct("2023-09-01 03:00:00", tz="UTC")
vardata = ib.cut %>% filter(assay_id %in% c(4, 11),
                                datetime > varstart,
                                datetime < varstop)
varsum = vardata %>% group_by(assay_id) %>%
  summarise(tsum = sum(abs(temp - TNEST)), tmean = mean(temp))
  varsum$y_manual = c(35, 33.75)

varexA = ggplot(vardata) +
  geom_line(aes(x = datetime, y = temp, color = assay_id)) +
  geom_hline(yintercept = TNEST, linetype = 2) +
  geom_ribbon(data = vardata %>% filter(assay_id == 4),  
              aes(ymax = TNEST, ymin = temp, xmin = varstart, 
                  xmax = varstop, x = datetime),
              fill = 'red', alpha = 0.3) +
  geom_ribbon(data = vardata %>% filter(assay_id == 11),  
              aes(ymax = TNEST, ymin = temp, xmin = varstart, 
                  xmax = varstop, x = datetime),
              fill = 'turquoise2', alpha = 0.3)+
  geom_label(data = varsum, aes(x = varstart + hours(3), 
             label = round(tsum, 0), color = assay_id, y = y_manual)) +
  labs(y = "Temperature (\u00B0C)", x = NULL) +
  theme(legend.position = "none") +
  theme(axis.text.x = element_blank())

#examples of variance 
hours = data.frame( datetime = 
  seq(
    from=varstart,
    to=varstop,
    by="hour"
  ))

exvar = ib.variables %>% filter(hour > varstart, 
                                hour < varstop + hours(1),
                                assay_id %in% c(4, 11)) %>%
  mutate(midhour = hour - minutes(30))

varexB = ggplot(vardata) +
  geom_line(aes(x = datetime, y = temp, color = assay_id)) +
  geom_vline(xintercept = hours$datetime, linetype = 2) +
  labs(y = "Temperature (\u00B0C)", x = NULL) +
  geom_label(data = exvar, aes(x = midhour, y = mtemp - 0.5, 
                                 label = round(var, 2), 
                                 color = assay_id)) +
  theme(legend.position = "none")

plot_grid(varexA, varexB, nrow = 2, labels = "AUTO")
#save figure
ggsave("figures/supplemental1_thermvars.pdf", device = "pdf", units = "in",
       width = 6.38, height = 6.54)
```


explicit pairing for t-tests
```{r}
#thermal error
pbar.err = ggplot(therm2, aes(x = cat, y = meanerr)) +
  #bars for means
  geom_col(data = therm.means, aes(y = mean_reperr, fill = rep),
           position = "dodge", alpha = 0.5, width = 0.5) + 
  #individual points
  geom_pointrange(aes(ymax = meanerr+seerr, ymin = meanerr-seerr, color = rep),
                  position=position_dodge(width=0.5),
                  show.legend = F) +
  #text and significance
  labs(y = "Hourly Error (\u00B0 C)" , fill = "replicate", x = NULL) +
  geom_signif( aes(color = NULL), step_increase = 0.1,
    comparisons = list(c("inbred", "hybrid"), c("inbred", "mixed"), c("hybrid", "mixed")),
    annotations = err.annotate, textsize = 4) +
  theme(axis.text.x.bottom = element_blank(),
        panel.border = element_blank(),
        axis.ticks.x = element_blank())


#plot thermal variance
pbar.var = ggplot(therm2, aes(x = cat, y = meanvar)) +
  #bars for means
  geom_col(data = therm.means, aes(y = mean_repvar, fill = rep),
           position = "dodge", alpha = 0.5, width = 0.5) + 
  #individual points
  geom_pointrange(aes(ymax = meanvar+sevar, ymin = meanvar-sevar, color = rep),
                  position=position_dodge(width=0.5),
                  show.legend = F) +
  #text and significance
  labs(y = bquote("Hourly Variance (\u00B0 C^2)"), fill = "replicate", x = NULL) +
  geom_signif( aes(color = NULL), step_increase = 0.1,
    comparisons = list(c("inbred", "hybrid"), c("inbred", "mixed"), c("hybrid", "mixed")),
    annotations = var.annotate, textsize = 4) +
  theme(panel.border = element_blank())

#supp figure
plot_grid(pbar.err, pbar.var, labels = "AUTO", nrow = 2)
#save figure
ggsave("figures/supplemental2_paired-t-test.pdf", device = "pdf", units = "in",
       width = 7.59, height = 6.54)

```

